### 2013-03-28 doc/sosp13/communication.tex
m

% In order to mitigate the low message injection rate limits of commodity
% networks, Grappa's communication stack has two layers:

@@

% When a task sends a message, the message is linked to a send queue
% associated with the message's destination node and the task continues
% execution. Note that the message contents stay in the worker's stack until
% it is sent to the network interface buffer, to avoid multiple copies of the
% payload.

@@

\paragraph{Message processing mechanics} Since communication is very frequent
in \Grappa, aggregating and sending messages efficiently is very important. To
achieve that, \Grappa makes careful use of caches, prefetching and lock-free
synchronization operations. \TODO{Does the explanation below need a figure?}

Outgoing messages are organized as an array of linked lists, with one linked
list per destination node in the system. Each core in a given node is
responsible for aggregating and sending the resulting messages to a set of
destination nodes. The outgoing message lists are located in a region of
memory shared across all cores in a \Grappa node (thus enabling cores to peek
at each other's message lists). When a task sends a message, it allocates a buffer (typically on its stack),  determines the destination system
node, and links the buffer into the corresponding linked list.

Each core has a system worker per destination node it is responsible for; this
system worker periodically checks whether messages to the corresponding
destination node is either large enough or has waited past a time-out period,
at which point the message is sent out. Aggregating and sending a message
involves manipulating a shared data-structure (the message list). This is done
using CAS (compare-and-swap) operations to avoid high synchronization costs.

% Threads that dispatch a message allocate a message buffer (typically on their
% stack) and dispatch a send request. \Grappa examines the destination system
% node for this field, and adds the message to a linked list of messages
% destined for that node. Periodically, or when this linked list grows large,
% the list is transferred using a CAS (compare-and-swap} operation to the
% processor on the originating system node that is responsible for managing all
% outbound communication requests to the specified destination node. This
% two-step process is used because it was determined empirically that we could
% only infrequently use processor coherent access requests because they remain a
% bottleneck on modern multicore hardware.

@@

Once the remote system node has received the message buffer, it is unpacked
and put in the corresponding incoming message lists. This process is done in
parallel by all system worker responsible for communication. Upper level
\Grappa messages are delivered to each core at the destination system node by
examining the message buffer at each processor core and then handing the
message buffer to the next core for delivery of messages destined to that
core.

>>>>

@@

@@

\paragraph{Message processing mechanics} Since communication is very frequent
in \Grappa, aggregating and sending messages efficiently is very important. To
achieve that, \Grappa makes careful use of caches, prefetching and lock-free
synchronization operations.

Each processing core of a system node maintains an array of outgoing message lists.  The array size is the number of system nodes in the \Grappa system.   The outgoing message lists and messages are located in a region of memory shared across all cores in a \Grappa node (thus enabling cores to peek at each other's message lists). When a task sends a message, it allocates a buffer (typically on its stack), determines the destination system node, and links the buffer into the corresponding linked list.

Each processing core in a given system node is responsible for aggregating and sending the resulting messages from all processors on that node to a set of destination nodes.  Each core periodically executes a task responsible for sending messages.  This task examines the private (to each core) message lists for each destination node it is responsible for managing and if the list is long enough, or a message has waited past a time-out period, all messages to a given destination system node from that source system node are sent.  Aggregating and sending a message involves manipulating a set of shared data-structures (the message lists). This is done using CAS (compare-and-swap) operations to avoid high synchronization costs.  Note that the reason we use a per-processor array of message lists that is only periodically modified across processor cores is we empirically determined this approach was faster (sometimes significantly) than a global per-system node array of message lists.

@@

Once the remote system node has received the message buffer, a management task is spawned to manage the unpacking process.  The management task spawns a task on each processor at the receiving system to simultaneously unpack messages destined for that processor.  Upon completion, these unpacking tasks synchronize with the management task.  Once all cores have processed the message buffer the management task sends a reply to the sending system node indicating the successful delivery of the messages.
